---
output:
  pdf_document: default
  html_document: default
  md_document:
    variant: markdown_github
---
# Basic Inferential Data Analysis

In the second portion of the project, we're going to analyze the ToothGrowth data in the R datasets package. From the R Documentation, ToothGrowth measures the effect of vitamin C on the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs.Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice (OJ) or ascorbic acid (a form of vitamin C and coded as VC).

## Load ToothGrowth Data and Basic Exploratory Data Analyses

```{r}
# load datasets and ggplot2 packages and ToothGrowth data
library(datasets); library(ggplot2)

data(ToothGrowth)

# examine structure of ToothGrowth
str(ToothGrowth)
```

`ToothGrowth` has 60 observations of 3 variables: length, supplement (OJ or VC), and dose (0.5, 1, 2).

For convenience, the dataset will be subsetted by supplement and dosage.
```{r}
# change dose to factor variable
ToothGrowth$dose <- as.factor(ToothGrowth$dose)

# subset length measurements by supplement and also dosage
ojlength <- subset(ToothGrowth, supp == 'OJ')$len
vclength <- subset(ToothGrowth, supp == 'VC')$len
length05 <- subset(ToothGrowth, dose == 0.5)$len
length1 <- subset(ToothGrowth, dose == 1)$len
length2 <- subset(ToothGrowth, dose == 2)$len
```

## Basic Summary of Data
```{r}
table(ToothGrowth$dose,ToothGrowth$supp)
```
Each supplement was administered to 30 guinea pigs, and for each supplement, there are three subgroups of 10 guinea pigs who received the three dosage levels.

### Tooth Growth by Supplement

Provide summary of `len` by `OJ`.
```{r}
summary(ojlength)
```

Provide summary of `len` by `VC`.
```{r}
summary(vclength)
```

Plot a boxplot of `len` by `supp`.
```{r}
g <- ggplot(ToothGrowth)
g + geom_boxplot(aes(supp, len, fill = supp)) +
        labs(title = 'Boxplot of Tooth Growth by Supplement',
             x = 'Supplement', y = 'Length')
```

From the summaries and boxplot above, orange juice, on average, is a more effective delivery method for increasing tooth growth. Ascorbic acid has greater variability in length than OJ; which will be calculated below.
```{r}
tapply(ToothGrowth$len, ToothGrowth$supp, var)
```

### Tooth Growth by Dosage

Provide summary of `len` by `0.5`.
```{r}
summary(length05)
```

Provide summary of `len` by `1`.
```{r}
summary(length1)
```

Provide summary of `len` by `2`.
```{r}
summary(length2)
```

Plot a boxplot of `len` by `dose`.
```{r}
g <- ggplot(ToothGrowth)
g + geom_boxplot(aes(dose, len, fill = dose)) +
        labs(title = 'Boxplot of Tooth Growth by Dosage',
             x = 'Dosage', y = 'Length')
```

From the summaries and boxplot above, it shows that a higher dosage results in longer lengths of odontoblasts.

### Tooth Growth by Supplement and Dosage

Plot a boxplot of length, by supplement and dosage:
```{r}
g <- ggplot(ToothGrowth)
g + geom_boxplot(aes(dose, len, fill = dose)) +
        facet_grid(.~supp) +
        labs(title="Tooth Length of Guinea Pigs by Dosage and Supplement",
             x="Dosage", y="Length")
```

It appears that increasing the dosage from 1 mg/day to 2 mg/day for orange juice did not have as much of an effect on the mean length as other dosage increases. Furthermore, there does not seem to be much a difference in mean length between supplements at the 2 mg/day dosage level.

## Hypothesis Testing

### Tooth Growth by Supplement

Perform a T-test for a statistically significant difference in mean length between ascorbic acid and orange juice. The groups are not paired, and assume unequal variance; these are `FALSE` by default.
```{r}
t.test(vclength, ojlength)
```

Since the p-value, 0.06063, is greater than 0.05, we fail to reject the null hypothesis that the difference in mean length for acsorbic acid and orange juice is zero.

### Tooth Growth by Dose

For the following tests, the groups are not paired, and assume unequal variance.  
  
Perform T-tests for a statistically significant difference in mean length for each dosage.
```{r}
# 0.5 vs 1
t.test(length05, length1)

# 1 vs 2
t.test(length1, length2)

# 0.5 vs 2
t.test(length05, length2)
```

The p-values for all three tests were less than 0.05, thus we reject the null hypothesis that the difference in means is zero. In other words there is a statistically significant increase in length as dosage increases.

### Further Testing

As mentioned earlier, there did not seem to be much of a difference in mean length for 1 mg/day and 2 mg/day in the orange juice group. Nor did there seem to be a difference for the 2 mg dosage between each supplement group. These will be further tested below.

```{r}
ojlength1 <- subset(ToothGrowth, supp == 'OJ' & dose == 1)$len
ojlength2 <- subset(ToothGrowth, supp == 'OJ' & dose == 2)$len

# OJ: 1 vs 2
t.test(ojlength1, ojlength2)
```

Since the p-value was less than 0.05, we reject the null hypothesis that the difference in mean length between 1 mg/day and 2 mg/day dosage in the orange juice group is zero. It may be worth noting that the p-value is is close to 0.04, and the null hypothesis could be rejected for lower alpha levels. This is also reflected in the 95% confidence interval; the maximum is -0.1885575, which is somewhat close to zero.

```{r}
vclength2 <- subset(ToothGrowth, supp == 'VC' & dose == 2)$len

# 2 vs 2
t.test(ojlength2, vclength2)
```

The p-value is over 0.05, thus we fail to reject the null hypothesis. In other words, there is no statistically significant difference in mean length between orange juice and ascorbic acid at a 2 mg/day dosage level.

## Assumptions and Conclusions

The following were some assumptions made:

1. The sample of 60 guinea pigs was representative of the guinea pig population
2. The samples are random iid
3. The population distribution of each sample group is approximately normal or mound shaped and roughly symmetric.

Thus we can conclude:

1. Supplement type does not affect the mean of length of the odontoblasts.
2. An increase in dosage results in an increase in tooth growth.
3. At 2 mg/day dosage, both supplements perform similarly.